---
title: "Clustering longline fishing strategies"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Clustering longline fishing strategies}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  chunk_output_type: console
params:
  scenario: "sp_lat"
  outdir: ""      # "" = a temporary directory; set a path to keep the outputs
  data_path: ""   # "" = bundled mock data; set the path to a seafisheries RDS
  n_pc: 0         # 0 = use pca_method; integer >= 2 overrides the number of PCs kept
---

```{r knit_options, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(longlineTypology)

## Base-R build: dplyr, lubridate and here are not required.
list2env(params, envir = .GlobalEnv)

if (outdir == "") outdir <- file.path(tempdir(), "cluster_outputs")
dir.create(outdir, showWarnings = FALSE, recursive = TRUE)

stat_list <- list()
```

This document demonstrates clustering of operational longline data to identify
fishing typologies. By default it runs on a small bundled mock dataset; for a
real analysis, point `data_path` to the output of the `seafisheries`
processing pipeline (same format), e.g.

```{r render_example, eval = FALSE}
rmarkdown::render("clusters.Rmd",
                  params = list(scenario  = "sp_hbf",
                                data_path = "path/to/LL_EC_LF.RDS",
                                outdir    = "cluster_outputs/"))
```

Summary:

- Recomputes species fractions for the species used by the scenario
- Selects relevant variables for clustering and filters invalid hbf / length rows
- Performs PCA to reduce dimensionality (number of components from the species
  count, a variance threshold, or a manual override)
- Uses the gap statistic over bootstrap replicates to find candidate numbers of
  clusters K
- Runs kmeans on the full dataset for every candidate K and assigns clusters
- Saves a diagnostics object that downstream reports read (no dataset inside)

# Load dataset

```{r load_dataset}
if (data_path == "") {
	data_path <- system.file("extdata", "mockdata", "LL_mock.RDS",
							 package = "longlineTypology")
	use_mock  <- TRUE
} else {
	use_mock  <- FALSE
}

EC_clean <- readRDS(data_path)
stat_list$nrow_total <- nrow(EC_clean)
```

The mock data predates the current `seafisheries` output format: hooks between
floats are stored as `hbf` and the outlier flag as a logical. The chunk below
brings it to the expected format (`mean_hbf`, `hbf_outlier` = `"valid"` /
`"outlier"`). It does nothing on real data.

```{r mock_format}
if (use_mock) {
	names(EC_clean)[names(EC_clean) == "hbf"] <- "mean_hbf"
	EC_clean$hbf_outlier <- ifelse(!is.na(EC_clean$mean_hbf) & !EC_clean$hbf_outlier,
								   "valid", "outlier")
}
```

# Define clustering parameters

```{r parameter_values}
# PCA parameters
pca_method             <- "n_components" # "threshold" or "n_components"
pca_variance_threshold <- 0.7            # Used only when pca_method = "threshold"

# kmeans / gap statistic parameters (sized for the mock data; for the full
# dataset use e.g. sample_no = 20000, nstart = 30)
sample_no              <- 100   # Sample size per replicate for gap statistic
kmeans_set             <- 15    # Number of kmeans replicates
max_k                  <- 20    # Maximum number of clusters to test
iter_max               <- 1e6   # Maximum iterations for kmeans
nstart                 <- 10    # Number of random starts for kmeans
B_var                  <- 25    # Bootstrap replicates for gap statistic
ncores                 <- 1    # Number of cores used by fastClusGap (Unix only)
```

How the number of principal components is chosen:

- `n_pc > 0` (parameter): manual override, clamped to the number of PCs the
  scenario provides.
- `pca_method = "n_components"`: `nSpeciesComponents()` keeps one component
  fewer than the number of species in the scenario (the species fractions sum
  to 1, so one of them is redundant), with a minimum of 2.
- `pca_method = "threshold"`: keeps enough components to explain
  `pca_variance_threshold` of the variance.

# Select and prepare variables

`selectScenario()` turns the scenario string into column names. Tokens are
joined with `_`, e.g. `yba` (yellowfin, bigeye, albacore), `ybas` (+ skipjack),
`sp` (yellowfin, bigeye, albacore, others), `hbf`, `len`, `lat`, `lon`.

`recalcSpeciesFractions()` recomputes `total_n` and the fractions so that the
fractions of the species *used by the scenario* sum to 1 in every row.

```{r select_prepare_vars}
EC_clean <- recalcSpeciesFractions(EC_clean, scenario)

pca_select <- selectScenario(scenario)

uses_hbf <- any(c("mean_hbf", "hbf_cat") %in% pca_select)
uses_len <- "mean_len" %in% pca_select

if (uses_hbf) {
	stopifnot("hbf_outlier" %in% names(EC_clean))
	EC_clean <- EC_clean[EC_clean$hbf_outlier == "valid", , drop = FALSE]
}
if (uses_len) {
	stopifnot("len_source" %in% names(EC_clean))
	EC_clean <- EC_clean[EC_clean$len_source != "imputed", , drop = FALSE]
}
stat_list$nrow_after_validity_filter <- nrow(EC_clean)

cat("Variables selected for clustering:\n")
print(pca_select)
cat("Rows retained after hbf/len validity filtering:", nrow(EC_clean), "\n")

select_dat <- EC_clean[, pca_select]
pca_scaled <- pcaScale(select_dat, method = "zscore")
```

# Principal Component Analysis

```{r pca}
pca_res <- customPCA(pca_scaled,
					 variance_threshold = pca_variance_threshold,
					 print_it = FALSE)

max_pc <- ncol(pca_res$x)

if (n_pc > 0) {
	if (n_pc != round(n_pc)) stop("n_pc must be an integer, got ", n_pc)
	n_pc_used <- as.integer(n_pc)
	if (n_pc_used < 2) stop("n_pc must be >= 2, got ", n_pc_used)
	if (max_pc < 2) stop("Scenario '", scenario, "' yields only ", max_pc,
						 " PC(s); at least 2 are required")
	if (n_pc_used > max_pc) {
		warning("n_pc = ", n_pc_used, " exceeds the ", max_pc,
				" PCs available for scenario '", scenario,
				"'; clamping to ", max_pc)
		n_pc_used <- max_pc
	}
	pca_res$no_var <- n_pc_used
	cat("PCA retained", pca_res$no_var, "components (manual override, requested",
		n_pc, "of", max_pc, "available)\n\n")
} else if (pca_method == "n_components") {
	pca_res$no_var <- nSpeciesComponents(scenario, min_components = 2)
	cat("PCA retained", pca_res$no_var, "components\n\n")
} else {
	cat("PCA retained", pca_res$no_var, "components explaining >=",
		pca_variance_threshold * 100, "% of variance\n\n")
}

pca_full <- pca_res$x[, 1:pca_res$no_var]

# Filter to post-1990 data for clustering
indices_90  <- which(as.integer(format(EC_clean$ymd, "%Y")) >= 1990)
pca_full_90 <- pca_full[indices_90, ]

cat("Using", nrow(pca_full_90), "observations from 1990 onwards for clustering\n")
stat_list$nrow_clustering <- nrow(pca_full_90)
```

# Determine candidate numbers of clusters

Each replicate draws `sample_no` rows (with replacement) from the post-1990
data and selects K with the gap statistic. Every K selected by at least one
replicate is carried forward; the choice between them is left to the
diagnostic report.

```{r optim_k}
set.seed(123)
sample_indices <- lapply(1:kmeans_set, function(x) {
	sample(1:nrow(pca_full_90), size = sample_no, replace = TRUE)
})

cat("\nRunning", kmeans_set, "kmeans replicates on samples of",
	sample_no, "observations...\n")

kmeans_res <- list()
gap_stat   <- list()

for (i in 1:kmeans_set) {
	pca_sub <- pca_full_90[sample_indices[[i]], 1:pca_res$no_var]

	res <- customKmeans(pca_sub,
						max_k      = max_k,
						random_set = B_var,
						iter_max   = iter_max,
						nstart     = nstart,
						d.power    = 2,
						ncores     = ncores,
						print_it   = FALSE)

	kmeans_res[[i]] <- res$kmeans
	gap_stat[[i]]   <- res$gap_stat
}

# K selected by each replicate
cluster_counts <- data.frame(
	cluster_n = sapply(kmeans_res, function(x) length(unique(x$cluster))),
	replicate = 1:kmeans_set
)

freq_table <- table(cluster_counts$cluster_n)

# Store as data frame for use in diagnostic report
freq_df <- data.frame(
	n_clusters = as.numeric(names(freq_table)),
	freq       = as.numeric(freq_table)
)

cat("\nCluster number distribution across replicates:\n")
print(freq_table)
```

# Apply clustering

For each candidate K, kmeans is fitted on the full post-1990 data, and every
row (including pre-1990 rows) is assigned to the nearest centroid with
`assignClusters()`. Results are stored as `cluster_k<K>` columns.

```{r kmeans}
unique_ks <- sort(unique(cluster_counts$cluster_n))

kmeans_full <- list()

for (k in unique_ks) {
	cat("\nRunning kmeans on full dataset with K =", k, "\n")

	res_k <- kmeans(pca_full_90, centers = k, iter.max = iter_max, nstart = nstart)

	raw_cluster <- assignClusters(as.data.frame(pca_full),
								  as.data.frame(res_k$centers))

	col_name <- paste0("cluster_k", k)
	EC_clean[[col_name]] <- as.integer(raw_cluster)
	kmeans_full[[as.character(k)]] <- res_k
}

cat("\nAdded columns:", paste0("cluster_k", unique_ks, collapse = ", "), "\n")
```

## Cluster summary per K

```{r final_cluster_summary, results = "asis"}
diag_tables <- list()

for (k in unique_ks) {
	tmp <- EC_clean
	tmp$cluster <- tmp[[paste0("cluster_k", k)]]

	summary_k <- summaryClusters(tmp)

	cluster_sizes       <- table(tmp$cluster)
	summary_k$n_obs     <- as.numeric(cluster_sizes[match(summary_k$cluster, names(cluster_sizes))])
	summary_k$pct_data  <- summary_k$n_obs / nrow(tmp) * 100

	diag_tables[[as.character(k)]] <- summary_k

	replicate_support <- freq_df$freq[freq_df$n_clusters == k]
	if (length(replicate_support) == 0) replicate_support <- 0

	betweenss_ratio <- kmeans_full[[as.character(k)]]$betweenss /
		kmeans_full[[as.character(k)]]$totss

	cat("\n### K =", k, "\n\n")
	cat("Replicate support:", replicate_support, "/", kmeans_set, "  \n")
	cat("Between-cluster SS / total SS:", round(betweenss_ratio, 3), "\n\n")
	print(knitr::kable(
		summary_k[, c("cluster", "n_obs", "pct_data", "yft_n", "total_n",
					  "yft_catch", "yft_frac")],
		format    = "simple",
		col.names = c("Cluster", "N", "% of Data", "YFT Caught", "Total Catch",
					  "% of Total", "% YFT in Cluster"),
		digits    = c(0, 0, 1, 0, 0, 1, 1)
	))
	cat("\n")
}
```

# Save outputs

The diagnostics object holds everything a downstream report needs, without the
dataset itself: run metadata, the scaling and PCA parameters to project new
data, the kmeans fits per K, the gap-statistic replicates and the cluster
assignment per row.

```{r save_outputs}
scale_params <- attr(pca_scaled, "scale_params")

# pca_res$x is the full per-row score matrix, same size as the dataset,
# and reconstructable via predict(pca_res_light, newdata = <new pca_scaled>),
# which only needs rotation/center/scale, not $x.
pca_res_light <- pca_res
pca_res_light$x <- NULL

diagnostics <- list(

	# run metadata
	scenario     = scenario,
	pca_method   = pca_method,
	pca_variance_threshold = if (pca_method == "threshold") pca_variance_threshold else NA,
	pca_select   = pca_select,      # variable names -- don't recover via rotation rownames, fragile
	n_components = pca_res$no_var,  # PCs actually used (after any clamping)
	n_pc_requested = n_pc,          # 0 = no override; otherwise the requested value
	n_pc_available = max_pc,

	# training data footprint: lets a future run check "does new data still
	# look like what this was fit on" without opening the raw file
	train_nrow       = nrow(EC_clean),
	train_nrow_1990  = nrow(pca_full_90),
	train_date_range = range(EC_clean$ymd),

	# apply the same PCA to new/future data
	scale_params = scale_params,    # mean/sd per raw variable
	pca_res      = pca_res_light,   # rotation + center + scale, for predict()

	# cluster identity -- assign new data to these same clusters later
	kmeans_full  = kmeans_full,     # named list keyed by K; centers live in each element

	# reproduce the gap statistic / centroid variability / silhouette / Dunn
	gap_stat       = gap_stat,      # per-replicate gap tables -> gap statistic plot
	kmeans_res     = kmeans_res,    # per-replicate kmeans fits -> centroid variability
	sample_indices = sample_indices,# exact rows sampled per replicate -> silhouette/Dunn need the *same* subsample
	freq_df        = freq_df,       # K frequency across replicates

	# cluster assignment, without the dataset itself
	cluster_k = EC_clean[, grep("^cluster_k", names(EC_clean)), drop = FALSE]
)

saveRDS(diagnostics, file.path(outdir, "cluster_diagnostics.RDS"))
cat("\nSaved diagnostics to", file.path(outdir, "cluster_diagnostics.RDS"), "\n")
```
